---
title: "Rank Compilation (across all years)"
output: html_notebook
---

I will be first compiling the ranks across the years. Here I will also choose to describe rank as dominance since normDS rather than ordinal ranks are used.

# Loading world environment
```{r}
rm(list = ls())

library(tidyverse)
```

# Loading required files
```{r}
rank_2026 <- read.csv("~/Savolainen Lab/SSB/Data/Rank/tidy_combined_rank_2026.csv") 

rank_2025 <- read.csv("~/Savolainen Lab/SSB/Data/Rank/Rank_per_year(in).csv")
```

# Merging both files
```{r}
rank_all <- bind_rows(rank_2026, rank_2025) %>% 
  arrange(Year)

write.csv(rank_all, "~/Savolainen Lab/SSB/Data/Rank/rank_master_combined_to2026.csv")
```

# Checking number of overlaps across the years
```{r}
repeated_monkeys <- rank_all %>%
  group_by(ID) %>%
  summarise(
    n_years = n_distinct(Year),
    years = paste(sort(unique(Year)), collapse = ", ")
  ) %>%
  filter(n_years > 1) %>%
  arrange(desc(n_years))

repeated_monkeys
```
# Preparing specific datasets
Here, I will only extract the relevant DS for each type of analysis

## DS for 2026 (for personality analyses)
I will use the original 2026 file and make some changes to the column for easy merging later on. This is for personality studies for only the 2026 monkeys.
```{r}
rank_2026_save <- rank_2026 %>% 
  rename(Animal_ID = ID,
         Group = Group_for_DS)

write.csv(rank_2026_save, "~/Savolainen Lab/SSB/Data/Rank/rank_2026_personality.csv")
```


## normDS for focal year (personality analyses)
Here I will extract the DS for the focal follows, based on the years the follow was carried out
```{r}
focals <- read.csv("~/Savolainen Lab/SSB/Data/Personality/Updated_Chloe/focals_used_forrank.csv")

rank_all_personality <- rank_all %>%
  mutate(
    ID = dplyr::recode(
      ID,
      "6.00E+06" = "6EE6",
      "5.00E+01" = "5EE1"
    )
  )

rank_personality <- focals %>% 
  left_join(
    rank_all_personality %>% 
      select(ID, Year, DS, DS_norm, Group_for_DS),
    by = c("ID", "Year")
  )

# 93R's rank is missing in 2024; Get 93R's 2023 rank values
rank_93R_2023 <- rank_all_personality %>%
  filter(ID == "93R", Year == 2023)

# Replace the missing 2024 values with the 2023 values
rank_personality <- rank_personality %>%
  mutate(
    DS = if_else(
      ID == "93R" & Year == 2024 & is.na(DS),
      rank_93R_2023$DS,
      DS
    ),
    DS_norm = if_else(
      ID == "93R" & Year == 2024 & is.na(DS_norm),
      rank_93R_2023$DS_norm,
      DS_norm
    ),
    Group_for_DS = if_else(
      ID == "93R" & Year == 2024 & is.na(Group_for_DS),
      rank_93R_2023$Group_for_DS,
      Group_for_DS
    )
  )

# 6EE6's rank is missing in 2026; Get 6EE6's 2025 rank values
rank_6E6_2025 <- rank_all_personality %>%
  filter(ID == "6EE6", Year == 2025)

# Replace the missing 2026 values with the 2025 values
rank_personality <- rank_personality %>%
  mutate(
    DS = if_else(
      ID == "6EE6" & Year == 2026 & is.na(DS),
      rank_6E6_2025$DS,
      DS
    ),
    DS_norm = if_else(
      ID == "6EE6" & Year == 2026 & is.na(DS_norm),
      rank_6E6_2025$DS_norm,
      DS_norm
    ),
    Group_for_DS = if_else(
      ID == "6EE6" & Year == 2026 & is.na(Group_for_DS),
      rank_6E6_2025$Group_for_DS,
      Group_for_DS
    )
  )

rank_personality <- rank_personality %>% 
  rename(Animal_ID = ID,
         Group = Group_for_DS)

write.csv(rank_personality, "~/Savolainen Lab/SSB/Data/Rank/rank_all_personality.csv")

```

## DS for 2023 (for gut analyses)
Here, 
```{r}
rank_gut <- rank_all %>% 
  filter(Year < 2024) %>% 
  arrange(ID, desc(Year)) %>%
  group_by(ID) %>%
  slice(1) %>%
  ungroup()

rank_gut <- rank_gut %>% 
  rename(Animal_ID = ID,
         Group = Group_for_DS)

write.csv(rank_gut, "~/Savolainen Lab/SSB/Data/Rank/rank_all_gut.csv")

```

## Mean DS Checks (for lifetime analyses)
Variance of normDS was checked and it was an average of 9.57 over 7 years (relatively stable for dominance ranks), yearly correlation is 0.887 (rather high). Graph looks concerning but overall seems fine, we can use mean DS

### Checking variance of DS
This checks if using a mean DS is a reasonable option
```{r}
rank_stability <- rank_all_personality %>%
  group_by(ID) %>%
  summarise(
    n_years = n(),
    mean_normDS = mean(DS_norm, na.rm = TRUE),
    sd_normDS = sd(DS_norm, na.rm = TRUE),
    min_normDS = min(DS_norm, na.rm = TRUE),
    max_normDS = max(DS_norm, na.rm = TRUE)
  ) %>%
  arrange(desc(sd_normDS))

stability_overall <- rank_stability %>%
  summarise(
    mean_SD = mean(sd_normDS, na.rm = TRUE),
    median_SD = median(sd_normDS, na.rm = TRUE)
  )
stability_overall

rank_stability
```

### Checking graphs
```{r}
ggplot(rank_all_personality,
       aes(x = Year,
           y = DS_norm,
           group = ID)) +
  geom_line(alpha = 0.4) +
  geom_point() +
  theme_bw()
```
### Yearly correlation
```{r}
rank_wide <- rank_all_personality %>%
  select(ID, Year, DS_norm) %>%
  pivot_wider(
    names_from = Year,
    values_from = DS_norm
  )

cor(rank_wide$`2023`,
    rank_wide$`2024`,
    use = "complete.obs")
```


## normDS mean (for lifetime analyses)
```{r}
mean_normDS <- rank_all_personality %>%
  filter(!is.na(DS_norm)) %>%
  group_by(ID) %>%
  summarise(
    mean_normDS = mean(DS_norm),
    years_normDS = n(),
    sd_normDS = sd(DS_norm),
    Group = last(Group_for_DS),
    .groups = "drop"
  )

mean_normDS <- mean_normDS %>% 
  rename(Animal_ID = ID)

write.csv(mean_normDS, "~/Savolainen Lab/SSB/Data/Rank/rank_mean_all.csv")
```

